---
title: "How to use a SQLite database from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use a SQLite database from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

```



## Load the libraries needed and set tha path to the database

The `DBI` package contains the functions needed, `RSQLite` the connectivity to an SQLite database. Sidenote: it is possible to switch to e.g. a MySQL database by simply using the appropriate package instead of `RSQLite`.

```{r}
library(DBI)
library(RSQLite)

dbname <- "~/LEEF/9.backend/LEEF.RRD.sqlite"  
```

The `::` are not necessary, but they make clear from which package the commands are
coming from and avoid using wrong functions from other packages.

I will only show the basic commands, a more detailed intro can be found at e.g. [datacamp](https://www.datacamp.com/community/tutorials/sqlite-in-r) and another basic intro at [RStudio](https://db.rstudio.com/databases/sqlite/).

## connect to database
```{r}
conn <- DBI::dbConnect(
	    		drv = RSQLite::SQLite(),
	    		dbname = dbname
	  		)
```
	  		
## List tables
```{r}
DBI::dbListTables(conn)
```
	
## List columns in tables
```{r}
DBI::dbListFields( conn, "flowcytometer" )
```

## Load complete table
```{r}
tbl <- DBI::dbReadTable( conn, "flowcytometer")
head(tbl)
```
	  		
## To the interesting stuff
In contrast to e.g. `.csv` files, databases can do joins, filter, calculations, etc. in the database without having to load all data into R. For this, the [SQLite (Sequential Query Language)](https://www.sqlitetutorial.net) languate is used. The basics are the same for all many databases (well - most SQL databases anyway). Please see the [SQLite tutorial for more info](https://www.sqlitetutorial.net).

### Read a table via SQL

```{r}
sql <- " SELECT * FROM flowcytometer"

tbl <- DBI::dbGetQuery(
  conn,
  statement = sql
)
head(tbl)
```

### Read only results from bottle 29

```{r}
sql <- " SELECT * FROM flowcytometer WHERE bottle = 29"

tbl <- DBI::dbGetQuery(
  conn,
  statement = sql
)
head(tbl)
```


### Read only selected columns from bottle 29

```{r}
sql <- " SELECT bottle, 'total.counts', gated_density_perml FROM flowcytometer WHERE bottle = 29"

tbl <- DBI::dbGetQuery(
  conn,
  statement = sql
)
head(tbl)
```


### join the manualcount into the flowcytometer

```{r}
sql <- " SELECT flowcytometer.bottle, flowcytometer.gated_density_perml, manualcount.density FROM flowcytometer INNER JOIN manualcount ON flowcytometer.bottle = manualcount.bottle"

tbl <- DBI::dbGetQuery(
  conn,
  statement = sql
)
head(tbl)
```

The sql statement can (and should now) be better not in a line:

```{r}
sql <- " 
SELECT 
  flowcytometer.bottle, 
  flowcytometer.gated_density_perml, 
  manualcount.density 
FROM 
  flowcytometer 
INNER JOIN 
  manualcount 
ON 
  flowcytometer.bottle = manualcount.bottle
"
```

which is much easier to read.

## Further points:

### [Indices](https://www.sqlitetutorial.net/sqlite-index/)
Databases use [Indices](https://www.sqlitetutorial.net/sqlite-index/) to make searching considerably faster than without indexes. I am building at
the moment indices from `timestamp, `bottle` (used for joining) and `timestamp x bottle`. If there are any other indices you think would be usefil, pleas let me know. No problem to add, even at a later stage.

### [Views](https://www.sqlitetutorial.net/sqlite-create-view/)
In addition to tables, which contain the data, databases have so called [Views](https://www.sqlitetutorial.net/sqlite-create-view/). Views is essentially a quer stored in the database, which can be read as a normal table. I am thinking of creating views which join the databases, contain common filters, etc. Please let me know whci ones would be useful for you.

### [Functions](https://www.sqlitetutorial.net/sqlite-functions/)
SQLIte includes a number of useful functions. Especially the [aggregate](https://www.sqlitetutorial.net/sqlite-aggregate-functions/) functions are very useful as, again, not all datapoints need to be read in R to be aggregated. This can be done by the database, using optimised code. We should discuss if this would be useful for you, to move some calculations into the database.


### Combining the different sql elements
All of the elements can be combined - you can have `SELECT` queries ( [subqueries](https://www.sqlitetutorial.net/sqlite-subquery/) ), which provide input to a function, which is also getting input from another subquery, and all of that is stored in a View. Many possibilities - and everything can be done from R.


## disconnect from database
This is important!

```{r}
DBI::dbDisconnect(conn)
```
